---
title: "Stuff"
author: "Franco C"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
hist=read_excel("C:\\Users\\franc\\OneDrive\\Hawks25\\NECBLHISTORY_combined.xlsx")
```



```{r}
# Load necessary libraries
library(dplyr)
library(caret)


hist <- hist %>%
  filter(!is.na(AutoPitchType), 
         AutoPitchType != "Other", 
         PitchCall != "Undefined", 
         PitchCall != "BallIntentional", 
         PitcherThrows != "Both", 
         BatterSide != "Undefined",
         !is.na(SpinRate))


hist <- hist %>%
  mutate(
      PitchCall = case_when(
      PitchCall %in% c("StikeCalled", "Strikecalled") ~ "StrikeCalled",
      PitchCall == "BallinDirt" ~ "BallCalled",
      PitchCall %in% c("FoulBall", "FoulBallNotFieldable", "FoulBallFieldable") ~ "Foul",
      PitchCall == "InPlay" & PlayResult %in% c("Single", "Double", "Triple", "HomeRun") ~ PlayResult,
      PitchCall == "InPlay" ~ "Out",
      TRUE ~ PitchCall
    ),
    BatterSide = ifelse(BatterSide == "RIght", "Right", BatterSide),
    PitcherThrows = ifelse(PitcherThrows == "RIght", "Right", PitcherThrows)
  ) %>%
  mutate(
    Balls = factor(Balls, levels = c(0,1,2,3), ordered = TRUE),
    Strikes = factor(Strikes, levels = c(0,1,2), ordered = TRUE),
    PitchCall = factor(PitchCall),
    PitcherThrows = factor(PitcherThrows, levels = c("Left", "Right")),
    BatterSide = factor(BatterSide, levels = c("Left", "Right"))
  )

```


```{r}
library(caret)
set.seed(2425)  


train_index <- createDataPartition(hist$PitchCall, p = 0.7, list = FALSE) 
train <- hist[train_index, ]
vali <- hist[-train_index, ]

```

```{r}
set.seed(2425)
library(randomForest)

rf_model <- randomForest(
  PitchCall ~ PitcherThrows+BatterSide+Balls+Strikes+RelSpeed+SpinRate+InducedVertBreak+HorzBreak+PlateLocHeight+PlateLocSide,
  data = train,  
  ntree = 100,  
  mtry = sqrt(10), #three vars randomly chosen at each node split in tree
  importance = TRUE,
)

print(rf_model)
```
misclassify 45% of pitches (OOB). Not terrible considering the # of classes
class.error is misclassification per class. 11% of balls misclassified. Since so many classes expect significant errors. Probability calibration is crucial


```{r}
varImpPlot(rf_model, main = "Feature Importance in Random Forest")

rf_pred <- predict(rf_model, newdata = vali)

# Compute Confusion Matrix
conf_matrix <- confusionMatrix(rf_pred, vali$PitchCall)
print(conf_matrix)

#some variables that are seemingly useless in the VIP should intuitively be kept. Pitcher and batter dexterity for example. Or count. 
#Overall accuracy is good relative to NIR. Significant improvement. 
#Sensitivity for in play outcomes are really poor. But I'm thinking maybe that makes sense since events like singles or doubles are extremely rare when considering all pitches. I think balls are plurality class at 38.72%. Since those events are so rare and we are using majority voting it makes sense that model would almost never give a single or double a plurality probability. 
```
```{r}
rf_prob <- predict(rf_model, newdata = vali, type = "prob")

y_actual <- model.matrix(~ PitchCall - 1, data = vali)

brier_score <- mean(rowSums((rf_prob - y_actual)^2) / ncol(y_actual))
cat("Brier Score:", brier_score, "\n")
#good
```


```{r}
preds <- as.data.frame(rf_prob)

preds$Actual <- vali$PitchCall #actual outcome
preds$Predicted <- predict(rf_model, newdata = vali) #predicted class (most likely)
preds$Strike=preds$StrikeCalled+preds$StrikeSwinging
preds <- cbind(preds, vali[, c("PitcherThrows", "BatterSide", "Balls", "Strikes", "AutoPitchType", "RelSpeed", "SpinRate", "InducedVertBreak", "HorzBreak", "PlateLocHeight", "PlateLocSide")])

View(preds)
```




```{r}
preds <- preds %>%
  mutate(
    Pitcher = vali$Pitcher,
    ActualRunValue = case_when(
      Strikes %in% c(0, 1) & Actual == "BallCalled" ~ 0.056,
      Strikes %in% c(0, 1) & Actual %in% c("StrikeCalled", "StrikeSwinging", "Foul") ~ -0.089,
      Strikes == 2 & Actual == "BallCalled" ~ 0.056,
      Strikes == 2 & Actual %in% c("StrikeCalled", "StrikeSwinging") ~ -0.089, #change to -0.26 since it is an out?
      Actual == "Out"        ~ -0.26,
      Actual == "Single"     ~ 0.44,
      Actual == "Double"     ~ 0.75,
      Actual == "Triple"     ~ 1.01,
      Actual == "HomeRun"    ~ 1.40,
      Actual == "HitByPitch" ~ 0.31,
      TRUE ~ 0  # Just in case there's anything unaccounted for
    )
  )
#estimated average run value of each event (MLB)
```



```{r}
# Run value weights
# Out = -0.25, Single = 0.47, Double = 0.78, Triple = 1.09, HR = 1.40

preds <- preds %>%
  mutate(
    xRunValue = case_when(
      Strikes %in% c(0, 1) ~ 
        0.056 * BallCalled -
        0.089 * (StrikeCalled + StrikeSwinging + Foul) +
        0.44  * Single +
        0.75  * Double +
        1.01  * Triple +
        1.40  * HomeRun -
        0.26 * Out +
        0.31  * HitByPitch,

      Strikes == 2 ~ 
        0.056 * BallCalled -
        0.089 * (StrikeCalled + StrikeSwinging) +
        0.44  * Single +
        0.75  * Double +
        1.01  * Triple +
        1.40  * HomeRun -
        0.26 * Out +
        0.31  * HitByPitch
    )
  )
#expected run value of each pitch using model probabilities
```

```{r}
pitcher_summary <- preds %>%
  group_by(Pitcher) %>%
  summarise(
    avg_xRunValue = mean(xRunValue, na.rm = TRUE), #average expected run value for all pitches
    avg_ActualRunValue = mean(ActualRunValue, na.rm = TRUE), #average actual run value based on outcomes
    pitches = n()
  ) %>%
  mutate(
    avg_xRunValue = round(avg_xRunValue, 5),
    avg_ActualRunValue = round(avg_ActualRunValue, 5),
    difference = round(avg_ActualRunValue - avg_xRunValue, 5)
  ) %>%
  arrange(difference)
#average actual run value and expected for each pitcher

```
```{r}
sum(pitcher_summary$avg_xRunValue)
sum(pitcher_summary$avg_ActualRunValue)
mean(pitcher_summary$avg_xRunValue)
mean(pitcher_summary$avg_ActualRunValue)
sum(preds$ActualRunValue)
sum(preds$xRunValue)
mean(preds$ActualRunValue)
mean(preds$xRunValue)
#elite validation performance. sums and means of expected vs actual are almost identical
```


```{r}
# One-hot encode actual classes
y_true <- model.matrix(~ PitchCall - 1, data = vali)

# Compute Brier score per class
brier_per_class <- colMeans((rf_prob - y_true)^2)

# Clean output
brier_per_class <- round(brier_per_class, 5)
print(brier_per_class)

#most are good, some are decent (foul)
```

```{r}
# Load and prepare
library(readr)
library(dplyr)
library(stringr)


# Load data
hist24 <- read_excel("C:/Users/franc/OneDrive/Hawks25/NECBLHISTORY_combined.xlsx")

# Remove August games and format pitcher name
hist24 <- hist24 %>%
   filter(
    substr(Date, 1, 4) == "2024",
    substr(Date, 6, 7) != "08"
  ) %>%
  mutate(
    PitcherFormatted = str_to_upper(sub("^(.*),\\s*(\\w).*", "\\1, \\2", Pitcher)),
    PitcherFormatted = str_replace_all(PitcherFormatted, "[‘’'`]", ""),
    PitcherTeam = str_trim(PitcherTeam)
  )

```

```{r}
hist24 <- hist24 %>%
  mutate(
    PlayResult = case_when(
      PlayResult %in% c("Error", "FieldersChoice", "Sacrifice") ~ "Out",
      TRUE ~ PlayResult
    )
  )

hist24 <- hist24 %>%
  filter(
    !is.na(AutoPitchType),
    AutoPitchType != "Other",
    PitchCall != "Undefined",
    PitchCall != "BallIntentional",
    PitcherThrows != "Both",
    BatterSide != "Undefined",
    !is.na(SpinRate)
  ) %>%
  mutate(
    PitchCall = case_when(
      PitchCall %in% c("StikeCalled", "Strikecalled") ~ "StrikeCalled",
      PitchCall == "BallinDirt" ~ "BallCalled",
      PitchCall %in% c("FoulBall", "FoulBallNotFieldable", "FoulBallFieldable") ~ "Foul",
      PitchCall == "InPlay" & PlayResult %in% c("Single", "Double", "Triple", "HomeRun") ~ PlayResult,
      PitchCall == "InPlay" ~ "Out",
      TRUE ~ PitchCall
    ),
    BatterSide = ifelse(BatterSide == "RIght", "Right", BatterSide),
    PitcherThrows = ifelse(PitcherThrows == "RIght", "Right", PitcherThrows),
    Balls = factor(Balls, levels = c(0,1,2,3), ordered = TRUE),
    Strikes = factor(Strikes, levels = c(0,1,2), ordered = TRUE),
    PitchCall = factor(PitchCall),
    PitcherThrows = factor(PitcherThrows, levels = c("Left", "Right")),
    BatterSide = factor(BatterSide, levels = c("Left", "Right"))
  )

colSums(is.na(hist24))
```



```{r}
hist24 <- hist24 %>%
  mutate(
    ActualRunValue = case_when(
      Strikes %in% c(0, 1) & PitchCall == "BallCalled" ~ 0.056,
      Strikes %in% c(0, 1) & PitchCall %in% c("StrikeCalled", "StrikeSwinging", "Foul") ~ -0.089,
      Strikes == 2 & PitchCall == "BallCalled" ~ 0.056,
      Strikes == 2 & PitchCall %in% c("StrikeCalled", "StrikeSwinging") ~ -0.089,
      PitchCall == "Out"        ~ -0.26,
      PitchCall == "Single"     ~ 0.44,
      PitchCall == "Double"     ~ 0.75,
      PitchCall == "Triple"     ~ 1.01,
      PitchCall == "HomeRun"    ~ 1.40,
      PitchCall == "HitByPitch" ~ 0.31,
      TRUE ~ 0
    )
  )

```

```{r}
# Predict probabilities from RF model
rf_prob_2024 <- predict(rf_model, newdata = hist24, type = "prob")
rf_prob_2024 <- as.data.frame(rf_prob_2024)


# Bind to main dataset
hist24 <- bind_cols(hist24, rf_prob_2024)

```

```{r}
hist24 <- hist24 %>%
  mutate(
    xRunValue = case_when(
      Strikes %in% c(0, 1) ~ 
        0.056 * BallCalled -
        0.089 * (StrikeCalled + StrikeSwinging + Foul) +
        0.44  * Single +
        0.75  * Double +
        1.01  * Triple +
        1.40  * HomeRun -
        0.26  * Out +
        0.31  * HitByPitch,

      Strikes == 2 ~ 
        0.056 * BallCalled -
        0.089 * (StrikeCalled + StrikeSwinging) +
        0.44  * Single +
        0.75  * Double +
        1.01  * Triple +
        1.40  * HomeRun -
        0.26  * Out +
        0.31  * HitByPitch
    )
  )

```

```{r}
pitcher_summary_2024 <- hist24 %>%
  group_by(PitcherFormatted, PitcherTeam) %>%
  summarise(
    avg_xRunValue = mean(xRunValue, na.rm = TRUE),
    avg_ActualRunValue = mean(ActualRunValue, na.rm = TRUE),
    pitches = n(),
    .groups = "drop"
  ) %>%
  mutate(
    avg_xRunValue = round(avg_xRunValue, 5),
    avg_ActualRunValue = round(avg_ActualRunValue, 5),
    difference = round(avg_ActualRunValue - avg_xRunValue, 5)
  )


```

```{r}
mean(hist24$xRunValue)
mean(hist24$ActualRunValue)
mean(hist24$xRunValue) - mean(hist24$ActualRunValue)


mean(preds$xRunValue)
mean(preds$ActualRunValue)
mean(preds$xRunValue) - mean(preds$ActualRunValue)
#this is what is making me suicidal. The average actual runs "saved" per pitch (better way to think about it) is significantly more when just narrowed to 2024 vs 2021-2024 and the model is not adjusting to the 2024 difference. Interesting thing is that this trend continues into 2023 (-0.1 actual average) but the model picks up on this and is significantly more accurate. There is a significant difference between average and expected runs saved per pitch in 2024 ( t tests below). But when I look at 2023, 2022, and 2021, the difference is not significant which is good. The model's failure in 2024 makes me nervous about 2025. Maybe 2024 is just an anomaly and offensive output was much better than previous years or it is more of a systematic miss. But comfortable sending this into 2025 and if there is another 2024 like miss we can explore options to fix it (options below)
```
```{r}
sum(hist24$xRunValue)
sum(hist24$ActualRunValue)
sum(hist24$xRunValue) - sum(hist24$ActualRunValue)

sum(preds$xRunValue)
sum(preds$ActualRunValue)
sum(preds$xRunValue) - sum(preds$ActualRunValue)
```


```{r}
t.test(preds$xRunValue, hist24$xRunValue, 
       alternative = "two.sided", var.equal = FALSE)


t.test(hist24$xRunValue, hist24$ActualRunValue, 
       alternative = "two.sided", var.equal = FALSE)

t.test(preds$xRunValue, preds$ActualRunValue, 
       alternative = "two.sided", var.equal = FALSE)
#not a significant difference in expected runs in validation and 2024 datasets, but model miss for 2024 is confirmed with t test between expected and actual. 
```

```{r}
library(readr)
library(dplyr)
library(stringr)

# Load the pitcher stats
pitching2024 <- read_csv("C:/Users/franc/OneDrive/Hawks25/necbl_combined_pitching_stats.csv")

# Clean pitcher names and team names
pitching2024 <- pitching2024 %>%
  mutate(
    Player = str_to_upper(Player),
    Player = str_replace_all(Player, "[‘’'`]", ""),
    Player = str_squish(Player),
    Team = str_trim(Team)
  )

# Prepare model-based pitcher summary for merge
pitcher_summary_2024 <- pitcher_summary_2024 %>%
  rename(Player = PitcherFormatted, Team = PitcherTeam)

merged_pitchers_2024 <- pitching2024 %>%
  left_join(pitcher_summary_2024, by = c("Player", "Team"))

```

```{r}
write_csv(merged_pitchers_2024, "C:/Users/franc/OneDrive/Hawks25/necbl_combined_pitching_stats.csv")
write_csv(merged_pitchers_2024, "C:/Users/franc/OneDrive/Hawks25/LeagueDataDashboard/necbl_combined_pitching_stats.csv")


```

options if 2025 is bad: Option 2: Recalibration (Platt scaling / isotonic regression)
When to use: Prediction probabilities are systematically misaligned (e.g., too much mass on low-run outcomes).

✅ Use true outcomes from 2025 to refit a probability adjustment layer, without retraining the whole model.

You can apply PlattScaling() or caret::calibration() to map predicted probabilities → better-calibrated probabilities

Recalculate xRunValue using those new probabilities

This is especially helpful when the model ranks predictions well but the probability magnitudes are off.

Option 3: Partial Re-training or Fine-tuning
When to use: You’ve collected enough 2025 data (e.g., 10k+ pitches) and the model is consistently wrong.

✅ Re-train using:

All past years + early 2025

Weighted training: e.g., more weight on 2024–2025

Use caret::train() or xgboost::xgb.train() with sample weights:
